---
title: "gnomeR for IMPACT data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{gnomeR-vignette}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message=F,warning=F,
  fig.width = 8,
  fig.height = 4
)
```


# Installation

```{r setup,eval=F}
install.packages("devtools")
devtools::install_github("AxelitoMartin/gnomeR")
```


# Requirements

<i>gnomeR</i> requires the following packages - ComplexHeatmap, iClusterPlus, cluster (installed with gnomeR).

```{r}
library(gnomeR)
```

```{r,include=F}
library(knitr)
library(dplyr)
library(dtplyr)
library(tibble)
library(plotly)
```

# Introduction {.tabset .tabset-fade .tabset-pills}

<i>gnomeR</i> is a R package that aims to process and analyze genetic data from cBioportal. We include in this package the mutation, copy number alteration (CNA), fusion and clinical information of all publicly available data from cBioPortal.

## Mutations 

```{r}
as.tbl(mut) %>% select(Tumor_Sample_Barcode,Hugo_Symbol,Hugo_Symbol,Variant_Classification,Mutation_Status)
```

## CNA

```{r}
as.tbl(cna[1:5,1:5])
```

## Fusions

```{r}
as.tbl(fusion) %>% select(Tumor_Sample_Barcode,Hugo_Symbol,Fusion)
```

## Clinical 

```{r}
as.tbl(head(clin))
```


# MAF processing

MAF files are the standard file format for mutation information. Each line represents a single mutation mapped to a sample, a particular gene and specific effect. All these fields are required in order to properly process the file. The IMPACT platform sequences a set of targeted oncogenic genes that are on cBioportal. 

## Creating a binary matrix

We can create a binary matrix of genetic events from the files described above. If a patient has a mutation in gene X the entry will be marked with a 1, otherwise it will be a 0. This function has the following arguments:

- patients : a character vector of sample IDs to be matched in the MAF file. Default is NULL in which case all samples will be used.
- maf : the corresponding MAF file (required)
- mut.type : the mutation type to be used, options are "SOMATIC", "GERMLINE" or "ALL". Default is "SOMATIC"
- SNP.only : boolean specifying if only SNPs should be used. Default is F
- include.silent : boolean specifying if silent mutations should be included. Default is F (recommended)
- fusion : optional fusion file to be included. Fusions will be added as their own column with ".fus" suffix
- cna : optional copy number alteration file to be included. CNAs will be added as their column with ".Amp" suffix for amplifications and ".Del" suffix for deletions
- cna.relax : boolean specifying if shallow deletions and low level gains should be counted as deletions and amplifications respectively

This function will return a binary matrix of genetics events with patients as rows and columns as genes. Along with a list of patients that weren't found to have any events (if any).

```{r}
patients <- as.character(unique(mut$Tumor_Sample_Barcode[1:10000]))
bin.mut <- create.bin.matrix(patients = patients,maf = mut,mut.type = "SOMATIC",SNP.only = F,include.silent = F)
as.tbl(bin.mut$mut)
```

Similarly including fusions and CNAs:

```{r}
bin.mut <- create.bin.matrix(patients = patients,maf = mut,mut.type = "SOMATIC",SNP.only = F,include.silent = F, fusion = fusion, cna = cna)
as.tbl(bin.mut$mut)
```

## Visualizing genetics

We include a function to visualize summaries of the mutations in a given cohort:

```{r}
maf.summary(maf = mut %>% filter(Tumor_Sample_Barcode %in% patients),
            mut.type = "SOMATIC")
```

# Correlating genetics with outcome

## Binary outcome

The `gen.tab` function allows us to test for potential differences in genetic event frequencies using Fisher's exact test for unpaired data and the McNemar exact test for paired data. This function takes the following arguments:

- gen.dat : a binary matrix of genetic events with patients as rows and columns as genes
- outcome : a binary vector corresponding to the patients in the gen.dat argument
- filter : a value in [0,1) that will remove all genes that have a frequency below that threshold. Default is 0, including all
- paired : boolean specifying is the observations are paired. Default is false
- cont : boolean specifying if the outcome is continuous. Default is false (see following section for an example)
- rank : boolean specifying if the output table should be ranked by pvalue

```{r}
outcome <- as.character(clin$Sex[match(patients,clin$DMPID)])
rmv <- ifelse(outcome %in% c("Male","Female"),F,T)
outcome <- outcome[-which(rmv)]
gen.dat <- bin.mut$mut[-which(rmv),]
gen.tab(gen.dat = gen.dat,
        outcome = outcome,
        filter = 0.05,paired = F,cont = F,rank = T)
```


## Continuous outcome

Similarly we show here an example with a simulated continuous outcome:

```{r}
outcome <-  rnorm(n = nrow(gen.dat))
tab.out <- gen.tab(gen.dat = gen.dat,
        outcome = outcome,
        filter = 0.05,paired = F,cont = T,rank = T)
tab.out$fits
tab.out$vPlot
```


## Time to event outcome

We further include `uni.cox` for univariate survival analysis if time to event data is available. This function takes as inputs: 

- X : a data frame containing the covariates to be used
- surv.dat : a data frame containing the time and status information 
- surv.formula : survival formula of the form `Surv(time,status)~.` . Note that delayed entry is allowed of the form `Surv(time1,time2,status)~.`
- filter : a value in [0,1) that will remove all genes that have a frequency below that threshold. Default is 0, including all

```{r}
surv.dat <- clin %>%
  filter(DMPID %in% patients,!is.na(time)) %>%
  select(DMPID,time,status) %>% 
  mutate(time = as.numeric(as.character(time)),
    status = ifelse(status == "LIVING",0,1))
X <- bin.mut$mut[match(surv.dat$DMPID,rownames(bin.mut$mut)),]
uni.cox(X = X, surv.dat = surv.dat,surv.formula = Surv(time,status)~.,filter = 0.05)
```


# Advanced genetic visuals

## OncoPrints

OncoPrints are a convenient way to study comutation patterns in our cohort through the `plot_oncoPrint` function. 

```{r}
# plot_oncoPrint(bin.mut$mut)
```



## FACETs 

