---
title: "Visualizations"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Visualizations}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
  
```{r, include = FALSE}
knitr::opts_chunk$set(
collapse = TRUE,
comment = "#>", warning = F, message = F
)
```

```{r, warning=F, message =F}
library(gnomeR)
library(dplyr)
```

# Introduction 

In this article we will explore the different automatized plotting functions included in `gnomeR`. There are two main functions in `gnomeR` to visualize genomic data: 

- `maf.summary()` provides an extensive summary of the mutations found in the MAF file provided
- `plot_oncoPrint()` provides an OncoPrint that let's the user explore the comutation (or any genetic event and clinical factors) patterns in the cohort of interest


# Summarizing a MAF file

The `maf.summary()` function takes the following inputs:

- `maf` a MAF file
- `...` any arguments that belong to the `binmat` method

`maf.summary()` will return a list of plots:

- `p.class` summarise the variant classification counts across the MAF 
- `p.type` summarise the variant type counts across the MAF 
- `p.SNV` summarise the SNV class counts across the MAF 
- `p.patient.variant` summarise the variants per sample counts across the MAF 
- `p.variant.bp` summarise the variants classification distribution 
- `p.genes` summarise the most frequently mutated genes counts
- `p.variant.dist` summarise the variants classification distribution as a percent of total variants
- `p.variant.dist.bar` summarise the variants classification distribution as a percent of total variants in a single sample
- `p.SNV.dist` summarise the SNV distribution as a percent of total SNV
- `p.corr` summarise the correlation of genes that are most frequently mutated in a heatmap
- `p.comut` summarise the comutation of genes that are most frequently mutated in a heatmap

An example of a subset of these plots is shown below.

```{r}
plots_summary <- maf.summary(maf = mut)
plots_summary$p.genes
plots_summary$p.variant.dist
plots_summary$p.comut
```


# OncoPrints

OncoPrint is a way to visualize multiple genomic alteration events by heatmap provided through the `ComplexHeatmap` package. We implemented an altered version of this function in `gnomeR` through the `plot_oncoPrint()` function that enables the user to explore the co-occurences in genomic and clinical features selected.

The `plot_oncoPrint()` function takes as arguments:
- `gen.dat` a genetic dataframe from `binmat()`
- `clin.dat` a set of clinical factors for the patients listed in gen.dat
- `ordered` a numeric vector specifying which order the samples should be plotted in (optional)

## Genetics only 

In the first section we show how to use the output of the `binmat()` output to create an OncoPrint for a selected subset of genes. For consistency we will use the 10 genes used in the comutation plot above.

```{r,fig.width=10, fig.height=7}
genes <- c("ATM","BRAF","MLL2","APC","EGFR","MLL3","PIK3CA","KRAS","TP53")
dat.gen <- binmat(maf = mut, fusion = fusion, cna = cna)
dat.gen <- dat.gen %>% select(starts_with(genes))
plot_oncoPrint(gen.dat = dat.gen)
```

## Adding clinical information 

In this section we will show how the user can further add clinical information to this plot.

```{r,fig.width=10, fig.height=7}
genes <- c("ATM","BRAF","MLL2","APC","EGFR","MLL3","PIK3CA","KRAS","TP53")
dat.gen <- binmat(maf = mut, fusion = fusion, cna = cna)
dat.gen <- dat.gen %>% select(starts_with(genes))
clin.patients.dat <- clin.patients[match(
  abbreviate(rownames(dat.gen),strict = TRUE, minlength = 9),
  clin.patients$X.Patient.Identifier),] %>%
  rename(DMPID = X.Patient.Identifier, Smoker = Smoking.History) %>%
  select(DMPID, Sex,Smoker) %>%
  filter(!is.na(DMPID)) %>%
  distinct(DMPID,.keep_all = TRUE)
dat.gen <- dat.gen[match(clin.patients.dat$DMPID,
                         abbreviate(rownames(dat.gen),strict = TRUE,
                                    minlength = 9)),]
clin.patients.dat <- clin.patients.dat %>%
  tibble::column_to_rownames('DMPID')
rownames(dat.gen) <- rownames(clin.patients.dat)
plot_oncoPrint(gen.dat = dat.gen,
               clin.dat = clin.patients.dat)
```


## Missing information

A critical point of `gnomeR` is its ability to correct for the different platforms that have existed at MSKCC. We reflect this is in the OncoPrints by including black entries in the missing cells as shown below.

```{r,fig.width=10, fig.height=7}
genes <- c("ATM","BRAF","MLL2","APC","EGFR","MLL3","PIK3CA","KRAS","TP53","EPHA7")
dat.gen <- binmat(maf = mut, fusion = fusion, cna = cna)
dat.gen <- dat.gen %>% select(starts_with(genes))
plot_oncoPrint(gen.dat = dat.gen)
```


# Advanced visualization 

The copy-number alterations data we have covered up to now is a discrete estimation of the alterations that occured. There however exist more nuanced and accurate data for copy-number alterations observed in a tumor. In `gnomeR` we include an example of segmentation file and relevant functions from the [`facets`](https://www.ncbi.nlm.nih.gov/pmc/articles/PMC5027494/) package that provides an allele-specific copy number and clonal heterogeneity analysis tool for high-throughput DNA sequencing. We show in this section how to visualize this data:

```{r, message = F, warning = F}
set.seed(123)
samples <- as.character(unique(mut$Tumor_Sample_Barcode))[sample(1:length(unique(mut$Tumor_Sample_Barcode)), 100, replace=FALSE)]
p.heat <- facets.heatmap(seg = seg, patients = samples, min.purity = 0)
p.heat$p
```

In the heatmap above we show the estimate gains and losses over segments in each chromosomes (rows) and in each patient (columns). Now that the purity of the tumor sample is an important factor in order to get reliable data. Moreover we include argument `ordered` and `outcome` that enables the user to sort the heatmap by a specific feature of interest. We show an example below with a simulated feature of outcome:

```{r, message = F, warning = F}
outcome <- rnorm(length(samples))
p.heat <- facets.heatmap(seg = seg, patients = samples, min.purity = 0, outcome = outcome, ordered = order(outcome))
p.heat$p
```
